Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2020-05-14 manuscript/manuscript.tex
ms updates

We also wanted to test whether we see similar convergence correlations under
Gaussian stabilizing selection. In these simulations, rather than targeting a
particular $V_A$, we fix the trait mutation rate at $10^{-8}$ (thus region-wide
$\theta = 2000$). Like the exponential directional selection simulations, we
impose directional selection in the same direction across the two populations
(converge), different directions (diverge), and only in one population
(single). We also vary the type (gradual versus sudden) and magnitude of optima
shifts in the two populations. Overall, simulations show convergence
correlations for the sudden optima shifts in Supplementary Figure
\ref{suppfig:convergence-corrs-gss}. Importantly, optima shifts in the opposite
direction cause negative convergence correlations. We found that for slow
moving optima shifts, the convergence correlations are generally too weak to be
distinguished from zero reliably, (top row of Supplementary Material Figure
\ref{suppfig:convergence-corrs-gss}).

@@

In our analysis of the \textcite{Barghi2019-qy} data, we describe our statistic
$G(t)$ as a lower bound for two reasons: (1) the population is sequenced every
ten generations, meaning the temporal covariances between adjacent generations
cannot contribute to the numerator of $G(t)$ but contribute to the denominator,
and (2) the estimate of $G(t)$ ignores linked selection's contribution to the
per-generation variance in allele frequency change (this is the difference
between our $G(t)$ and $G'(t)$ estimators, the latter of which includes these
variance terms \parencite{Buffalo2019-io}).  To verify that $G(t)$ estimated
every ten generations is indeed a lower bound, we used a simulation procedure
similar to the exponential fitness function simulations (described in
Supplementary Material Section \ref{supp:genetic-arch}), and calculated the
temporal covariances and $G(t)$ both each generation, and every ten
generations. Unlike the simulations described in \ref{supp:genetic-arch}, we
began selection at $10N$ generations, and used trait $V_A = 0.01$ and targeted
$L = 1000$ sites affecting the trait.

First, comparing $G(t)$ when sampling population frequencies every generation
versus every ten generations, we confirm that the ten-generation block $G(t)$
is a lower bound of the $G(t)$ trajectory when sampling is every generation
(red and blue lines in Supplementary Figure \ref{suppfig:supp-blocks}A).
Furthermore, since we control the population size in our simulations at $N =
1000$ diploids, we know the drift effective population size in the absence of
selection. This allows us to estimate $G(t)'$, which is a measure of $G(t)$
that accounts for the linked selection's inflation of the variance in allele
frequency change between two generations (equation 26, \cite{Buffalo2019-io}).
Plugging in the drift effective population size $N_e = 1000$ into the
expression for $G'(t)$ and using the $\var(p_t - p_0)$ calculated for different
$t$'s, we see that the every generation $G(t)$ that does not account for linked
selection's inflation of $\var(\Delta p_t)$ does underestimate the true impact
of linked selection as expected (dashed gray line in Supplementary Figure
\ref{suppfig:supp-blocks}A).

@@

In our previous work, \textcite{Buffalo2019-io}, we did not investigate whether
background selection can lead to temporal autocovariance. Here, using
forward-in-time simulations, we find that background selection can indeed
generate temporal autocovariance and lead to convergence correlations when
deleterious haplotypes are shared between populations and both removed by
selection.

We simulated background selection in a 50 megabase region, where deleterious
alleles are randomly introduced by mutation. Following background selection
literature
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1994-oh,Hudson1995-xc}, we
parameterize the mutation rate as the total number of deleterious mutations
introduced per diploid genome, per generation, and simulate values $U = \{0.5,
  1.0, 1.5\}$. \vb{Note that for our 50Mb region, our choice of BGS $U$
  parameters are on the higher end of the spectrum expected for
  \emph{Drosophila}, but we wanted to ensure the strength was sufficient to see
  a signal. Note that if $U \approx 1.6$ \parencite{Elyashiv2016-vt}, and the
\emph{Drosophila} genome is $\approx 140$Mb, our region is $\approx 36\%$ of
the genome; this implies a reasonable $U$ for our region is $U \approx 0.57$,
which is close to the bottom end of our parameter range.} We also vary the
strength of selection against the deleterious mutations, $s = \{0.0, 0.01,
0.05, 0.1\}$ (where $s=0$ is a neutral control), as well as different
recombination rates ($r_{bp} = \{10^{-7}, 10^{-8}\}$).  Like other simulations,
we burnin the population for $10N$ generations under backgrounds selection.
Overall, we find background selection does create temporal covariance
(Supplementary Material Figure \ref{suppfig:supp-bgs-covs}), which are stronger
under (1) higher deleterious mutation rates and (2) larger selection
coefficients. This latter point initially seems at odds with background
selection theory, as the level of pairwise diversity in a region under strong
background selection is invariant with respect to the selection coefficient.
However, looking at the background selection $G(t)$ trajectories, we find that
over time, the impact of linked selection created by background selection
appears to trend towards an equilibrium in the $r_{BP} = 10^{-7}$ subfigure,
and reaches an equilibrium in the $r_{BP} = 10^{-8}$ subfigure that seems
reasonably invariant to the choice of $s$ (Supplementary Material Figure
\ref{suppfig:supp-bgs-g-fix}). We believe that these observations can be
reconciled by $Cov(\Delta p_t,~ \Delta p_{t^{\prime}})$ being larger for larger
$s$ when $|t^{\prime}-t|$ is small, but also decaying more rapidly with
$|t^{\prime}-t|$, such that the overall contribution of selection to allele
frequency variance is invariant to $s$ (for strong background selection).
However, further future work is needed to fully explore and understand the
temporal covariance dynamics of background selection.

@@

\vb{Finally, we found in these background selection results that including or
excluding sites fixed or lost through time can lead  to differences in the
estimated $G(t)$ trajectories and temporal covariance. We discuss this extensively in a
subsequent section, Supplementary Materials Section \ref{supp:fixation}.}

@@

We \vb{also} explored how directional truncation selection generates temporal
covariances. In these simulations we select the top \vb{10\%, 25\%, or 50\%} of
the phenotypic distribution of individuals to form the next generation. We
burnin these simulations \vb{using a neutral burnin routine, where trait
alleles are selectively neutral until directional selection is imposed}. More
extreme directional truncation selection generated larger initial covariances
(Figure \ref{suppfig:supp-trunc}B).  However, weaker truncation selection
generated more sustained positive covariances and so had a larger long-term
impact on the variance in allele frequency change.  We again noticed strong
effect fixation  or loss of sites has on temporal covariance and $G(t)$ under
truncation selection.  Here, since only a (potentially small) fraction of
individuals contribute to the next generation, sites can fix over very short
timescales. Furthermore, the small number of effective breeders contributing to
the next generation shrinks $N_e$ considerably, which increases $\var(p_t -
p_0)$, the denominator of $G(t)$. We see the effect of handling fixed/lost
sites differently, and the faster rate of drift in Supplementary Materials
Figure \ref{suppfig:supp-trunc}A, where weaker truncation selection actually
has higher levels of $G(t)$. Looking just at temporal covariances, we find that
stronger truncation selection (e.g.  a smaller tail of individuals selected)
does lead to greater temporal covariances. \vb{Overall, these truncation
  selection simulations demonstrate the value of considering both absolute
  measures of selection's effect on allele frequency changes, i.e. temporal
  covariance, as well as measures relative to drift, i.e. the $G(t)$
  trajectories. While selecting a smaller tail of individuals is associated
  with \emph{stronger} selection, it also is leading to a faster rate of drift,
which can distort conclusions inferred from considering the $G(t)$ trajectories
alone.}

@@

To ensure that our findings were robust to handling sites with a frequency of
zero or one differently, we regenerated Figures \ref{fig:figure-1} and Figure
\ref{fig:figure-3} but excluded frequencies of zero or one. Specifically, we
wanted to ensure that our finding that \vb{temporal covariances were negative
at later timepoints} was not spuriously caused by the way fixed or lost sites
are handled. We see no qualitative difference (Supplementary Material Figures
\ref{suppfig:supp-fig-1-nofix} and \ref{suppfig:supp-fig-3-nofix}) that emerges
when sites with frequency zero or one are excluded.

@@

We have investigated the effectiveness of our correction on real data by
exploiting the relationship between sampling depth and the magnitude of the
variance and covariance biases, and comparing the observed variances and
covariances before and after correction. We plot the variance and covariance
(between adjacent timepoints) before and after the bias correction against the
average sample depth in 100kb genomic windows in Figure
\ref{suppfig:barghi-correction}. Overall, we find the biased-correction
procedure removes the relationship between variance and covariance and depth, indicating it is working adequately.

@@

  \caption{The genome-wide covariance ($\cov(\Delta p_0, \Delta p_10)$ pooling
    all replicates) averaged (red line) and the median windowed covariance
    (blue) for the \textcite{Barghi2019-qy} dataset. The trimmed average window
    covariance, excluding the $\alpha$ lower and upper tails, converges to the
     median windowed covariance. This indicates that genome-wide covariance are
     not being overly dominated by a large-effect loci in few windows.}

>>>>

We also wanted to test whether we see similar convergence correlations under
Gaussian stabilizing selection. In these simulations, rather than targeting a
particular $V_A$, we fix the trait mutation rate at $10^{-8}$ (thus region-wide
$\theta = 2000$). Like the exponential directional selection simulations, we
impose directional selection in the same direction across the two populations
(converge), different directions (diverge), and only in one population
(single). We also vary the type (gradual versus sudden) and magnitude of optima
shifts in the two populations. Overall, simulations show convergence
correlations for the sudden optima shifts in Supplementary Figure
\ref{suppfig:convergence-corrs-gss}. Again, optima shifts in the opposite
direction cause negative convergence correlations. We found that for slow
moving optima shifts, the convergence correlations are generally too weak to be
distinguished from zero reliably, (top row of Supplementary Material Figure
\ref{suppfig:convergence-corrs-gss}).

@@

In our analysis of the \textcite{Barghi2019-qy} data, we describe our statistic
$G(t)$ as a lower bound for two reasons: (1) the population is sequenced every
ten generations, meaning the temporal covariances between adjacent generations
cannot contribute to the numerator of $G(t)$ but contributes to the denominator,
and (2) the estimate of $G(t)$ ignores linked selection's contribution to the
per-generation variance in allele frequency change (this is the difference
between our $G(t)$ and $G'(t)$ estimators, the latter of which includes these
variance terms \parencite{Buffalo2019-io}).  To verify that $G(t)$ estimated
every ten generations is indeed a lower bound, we used a simulation procedure
similar to the exponential fitness function simulations (described in
Supplementary Material Section \ref{supp:genetic-arch}), and calculated the
temporal covariances and $G(t)$ both each generation, and every ten
generations. Unlike the simulations described in \ref{supp:genetic-arch}, we
began selection at $10N$ generations, and used trait $V_A = 0.01$ and targeted
$L = 1000$ sites affecting the trait.

First, comparing $G(t)$ when sampling population frequencies every generation
versus every ten generations, we confirm that the ten-generation block $G(t)$
is a lower bound of the $G(t)$ trajectory when sampling is every generation
(red and blue lines in Supplementary Figure \ref{suppfig:supp-blocks}A).
Furthermore, since we control the population size in our simulations at $N =
1000$ diploids, we know the drift effective population size in the absence of
selection. This allows us to estimate $G(t)'$, which is a measure of $G(t)$
that accounts for the linked selection's inflation of the variance in allele
frequency change between two generations (equation 26, \cite{Buffalo2019-io}).
Plugging in the drift effective population size $N_e = 1000$ into the
expression for $G'(t)$ and using the $\var(p_t - p_0)$ calculated for different
$t$'s, we see that $G(t)$ \vb{calculated every generation} does not account for
linked selection's inflation of $\var(\Delta p_t)$ and underestimates the true
impact of linked selection as expected (dashed gray line in Supplementary
Figure \ref{suppfig:supp-blocks}A).

@@

In our previous work, \textcite{Buffalo2019-io}, we did not investigate whether
background selection can lead to temporal autocovariance. Here, using
forward-in-time simulations, we find that background selection can indeed
generate temporal autocovariance and lead to convergence correlations when
deleterious haplotypes are shared between populations and both are removed by
selection.

We simulated background selection in a 50 megabase region, where deleterious
alleles are randomly introduced by mutation. Following background selection
literature
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1994-oh,Hudson1995-xc}, we
parameterize the mutation rate as the total number of deleterious mutations
introduced per diploid genome, per generation, and simulate values $U = \{0.5,
  1.0, 1.5\}$. \vb{Note that for our 50Mb region, our choice of BGS $U$
  parameters are on the higher end of the spectrum expected for
  \emph{Drosophila}, but we wanted to ensure the strength was sufficient to see
  a signal. Note that if $U \approx 1.6$ \parencite{Elyashiv2016-vt}, and the
\emph{Drosophila} genome is $\approx 140$Mb, our region is $\approx 36\%$ of
the genome; this implies a reasonable $U$ for our region is $U \approx 0.57$,
which is close to the bottom end of our parameter range.} We also vary the
strength of selection against the deleterious mutations, $s = \{0.01, 0.05,
0.1\}$, as well as different recombination rates ($r_{bp} = \{10^{-7},
10^{-8}\}$).  Like other simulations, we burnin the population for $10N$
generations under backgrounds selection.  Overall, we find background selection
does create temporal covariance (Supplementary Material Figure
\ref{suppfig:supp-bgs-covs}), which are stronger under (1) higher deleterious
mutation rates and (2) larger selection coefficients. This latter point
initially seems at odds with background selection theory, as the level of
pairwise diversity in a region under strong background selection is invariant
with respect to the selection coefficient.  However, looking at the background
selection $G(t)$ trajectories, we find that over time, the impact of linked
selection created by background selection appears to trend towards an
equilibrium in the $r_{BP} = 10^{-7}$ subfigure, and reaches an equilibrium in
the $r_{BP} = 10^{-8}$ subfigure that seems reasonably invariant to the choice
of $s$ (Supplementary Material Figure \ref{suppfig:supp-bgs-g-fix}). We believe
that these observations can be reconciled by $Cov(\Delta p_t,~ \Delta
p_{t^{\prime}})$ being larger for larger $s$ when $|t^{\prime}-t|$ is small,
but also decaying more rapidly with $|t^{\prime}-t|$, such that the overall
contribution of selection to allele frequency variance is invariant to $s$ (for
strong background selection).  However, further future work is needed to fully
explore and understand the temporal covariance dynamics of background
selection.

@@

Finally, we found in these background selection results that including or
excluding sites fixed or lost through time can lead  to differences in the
estimated $G(t)$ trajectories and temporal covariance. We discuss this
extensively in a subsequent section, Supplementary Materials Section
\ref{supp:fixation}.

@@

We also explored how directional truncation selection generates temporal
covariances. In these simulations we select the top 10\%, 25\%, or 50\% of the
phenotypic distribution of individuals to form the next generation. We burnin
these simulations using a neutral burnin routine, where trait alleles are
selectively neutral until directional selection is imposed. More extreme
directional truncation selection generated larger initial covariances
(Supplementary Material Figure \ref{suppfig:supp-trunc}B).  However, weaker
truncation selection generated more sustained positive covariances and so had a
larger long-term impact on the variance in allele frequency change.  We again
noticed fixation or loss of sites has \vb{a strong effect} on temporal
covariance and $G(t)$ under truncation selection.  Here, since only a
(potentially small) fraction of individuals contribute to the next generation,
sites can fix over very short timescales.  Furthermore, the small number of
effective breeders contributing to the next generation shrinks $N_e$
considerably, which increases $\var(p_t - p_0)$, the denominator of $G(t)$. We
see \vb{both} the effect of handling fixed/lost sites differently and the
faster rate of drift in Supplementary Materials Figure
\ref{suppfig:supp-trunc}A, where weaker truncation selection actually has
higher levels of $G(t)$. Looking just at temporal covariances, we find that
stronger truncation selection (e.g.  a smaller tail of individuals selected)
does lead to greater temporal covariances. Overall, these truncation selection
simulations demonstrate the value of considering both absolute measures of
selection's effect on allele frequency changes, i.e. temporal covariance, as
well as measures relative to drift, i.e. the $G(t)$ trajectories. While
selecting a smaller tail of individuals is associated with \emph{stronger}
selection, it also is leading to a faster rate of drift, which can distort
conclusions inferred from considering the $G(t)$ trajectories alone.

@@

To ensure that our findings were robust to handling sites with a frequency of
zero or one differently, we regenerated Figures \ref{fig:figure-1} and Figure
\ref{fig:figure-3} but excluded frequencies of zero or one. Specifically, we
wanted to ensure that our finding that temporal covariances were negative at
later timepoints was not spuriously caused by the way fixed or lost sites are
handled. We see no qualitative difference (Supplementary Material Figures
\ref{suppfig:supp-fig-1-nofix} and \ref{suppfig:supp-fig-3-nofix}) that emerges
when sites with frequency zero or one are excluded.

@@

We have investigated the effectiveness of our correction on real data by
exploiting the relationship between sampling depth and the magnitude of the
variance and covariance biases, and comparing the observed variances and
covariances before and after correction. We plot the variance and covariance
(between adjacent time intervals) before and after the bias correction against
the average sample depth in 100kb genomic windows in Figure
\ref{suppfig:barghi-correction}. Overall, we find the biased-correction
procedure removes the relationship between variance and covariance and depth,
indicating it is working adequately.

@@

  \caption{The genome-wide covariance ($\cov(\Delta p_0, \Delta p_{10})$
    pooling all replicates) averaged (red line) and the median windowed
    covariance (blue) for the \textcite{Barghi2019-qy} dataset. The trimmed
    average window covariance, excluding the $\alpha$ lower and upper tails,
    converges to the median windowed covariance. This indicates that
  genome-wide covariances are not being overly dominated by a large-effect loci
in few windows.}
